---
title: "csv - read/write"
subtitle: "" 
author: ""
output:
  html_document:
    toc: true
    toc_float: true
    highlight: "tango"
editor_options: 
  chunk_output_type: console
---

## Read (Import)

### base R `read.csv()`

```{r}
read.csv("example.csv")
```

- output is data frame (base R)
- missing for character is blank
- missing for numeric is 'NA' (not colored)
- spaces or special characters in headers are replaced with period 
- ï.. added to beginning of first column name (to remove use `read.csv("example.csv", fileEncoding = 'UTF-8-BOM')`; [Source](https://www.roelpeters.be/removing-i-umlaut-two-dots-data-frame-column-read-csv/))
- remove headers with `header = FALSE`; column names will be V1, V2, V3, etc. 


### `readr` `read_csv()`

```{r}
library(readr) #included in tidyverse
read_csv("example.csv")
```

- output is tibble (tidyverse)
- receive message with type of columns that R is using for import 
- missing for character and numeric is NA (colored red) in R console; outputs (html) will show has <NA> for character and NA for numeric 
- headers are that have spaces or special characters are placed within backticks (``)
- remove headers with `col_names = FALSE`; column names will be X1, X2, X3, etc. 


## Specify Column Types 

R does a pretty good job of figuring out what the columns should be but if its needed to specify column types (or you want don't want the default col_types message to show) column types can be specified:  

```{r}
read_csv("example.csv"
         , col_types = cols(
            `Name 1`   = col_character()
          , `Name (2)` = col_double()
          , `Name #3`  = col_double()
           )
)
```


## Write (Export) 

Can use `write_csv()` or `write.csv()` - have sightly different functionality.    

```{r}
test <- read_csv("example.csv"
         , col_types = cols(
            `Name 1`   = col_character()
          , `Name (2)` = col_double()
          , `Name #3`  = col_double()
           )
)
write_csv(test, "example_export1.csv")
write.csv(test, "example_export2.csv")
```

Rownames; `row.names = TRUE` to include; `row.names = FALSE` to exclude 

* `write.csv()` default includes row names (usually row number) 
* `write_csv()` default does not include row names; CANNOT ADD

&nbsp; 

NA values; `na = ""` to have missing data be exported as blank cell 

* If data set is a base R data frame: 
  + `write.csv()` default is `na = "NA"` for numeric, always blank for character (can't change!) 
  + `write_csv()` default is `na = "NA"` for numeric, always blank for character (can't change!) 
* If data set is a tibble: 
  + `write.csv()` default is `na = "NA"` for numeric and character 
  + `write_csv()` default is `na = "NA"` for numeric and character 


