---
title: "ggplot"
subtitle: "" 
author: ""
output:
  html_document:
    toc: true
    toc_float: true
    highlight: "tango"
editor_options: 
  chunk_output_type: console
---

```{r, message=FALSE}
# load-packages
library(tidyverse)
```

## Online books
<a href="https://ggplot2-book.org/" target="_blank"> ggplot2 - Hadley Wickham </a>  

<a href="https://serialmentor.com/dataviz/" target="_blank"> Fundamentals of Data Visualization - Claus O. Wilke </a>

## hjust and vjust
Why do I always forget the direction of these? 

**hjust:** 0 = left-aligned, 0.5=center, 1 = right-aligned  
**vjust:** 0 = top-aligned, 0.5=middle, 1 = bottom-aligned  

<a href="https://www.r-bloggers.com/hjust-and-vjust/" target="_blank"> Visual Example - R-bloggers </a> 

## Math Expressions in labels 

### Use `quote()` 
```{r}
ggplot(mpg, aes(displ, hwy))+geom_point()+
  ggtitle(
    quote(
      alpha ^ 2 - frac(1, 10) + sum(n[i], i==1, N)
                )
    )
```

### Use `TeX()` from the `latex2exp` package  
- must be in a string
- must be denoted as math mode with dollar signs
- must include **2** backslashes for $\LaTeX$ commands 
```{r}
library(latex2exp)
ggplot(mpg, aes(displ, hwy))+geom_point()+
  ggtitle(TeX(
    "$\\alpha^2 - \\frac{1}{10} + \\sum_{i}^N n_i$"
                )
    )
```


## Laying out multiple plots 
<a href="https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html" target="_blank"> Laying out multiple plots on a page - Baptiste Auguie </a>  


## Line up axes on stacked plots 
Sometimes I'm working on two different types of plots (like a bar chart and a scatter plot) that happen to have the same x-axis.  I want to line up these axes so that when the plots are stacked the values correspond to the same date.  




### `gridExtra::grid.arrange()` and `cowplot::plot_grid()`
```{r}
# two different bar charts
A <- ggplot(mpg, aes(class))+geom_bar()+coord_flip()+ylim(0, 109)
B <- ggplot(mpg, aes(drv))+geom_bar()+coord_flip()+ylim(0, 109)
```

 Using `grid.arrange` command from the `gridExtra` package does not line up axes.  
```{r}
#axes don't line up
gridExtra::grid.arrange(A, B, ncol=1)
```

Use `grid.draw` command from the `grid` package.  
<a href="https://stackoverflow.com/questions/13294952/left-align-two-graph-edges-ggplot" target="_blank"> Source </a>
```{r}
#make plots into Grobs (grid graphical object)
gA <- ggplotGrob(A) 
gB <- ggplotGrob(B)
grid::grid.draw(rbind(gA, gB))
```

The `cowplot::plot_grid()` function allows you to line up plots by a specific axis. 

```{r}
cowplot::plot_grid(A, B, ncol = 1, align = "v")
```

### Facets 

Another option is `facet_wrap()` or `facet_grid()`, which can works  if the axes are the same for the different variables you want to compare, but be careful as facets are supposed to be comparing items with the same measurements.  

```{r}
tidy.df <- pivot_longer(mpg, c(class, drv), names_to = "category", values_to = "type")

ggplot(tidy.df, aes(type))+
  geom_bar()+
  coord_flip()+ 
  facet_wrap(
      ~category
    ,  ncol = 1
    , scales = "free" #removes types from the axis if that category has 0 cars of that type 
  )


ggplot(tidy.df, aes(type))+
  geom_bar()+
  coord_flip()+ 
  facet_grid(
      category ~ . 
    , scales = "free" #removes types from the axis if that category has 0 cars of that type 
    , space  = "free" #spaces based on number of obs (i.e. number of bars);
                      #         rather than giving each facet equal sizing
  )
```



### Mixed Geoms (Bar + Scatter)  
Scatter plots and bar charts will not line up automatically, even when using the `grid.draw` command detailed above.  This is because their default limits are different given that the bar chart is centered on the value and the scatter plot is a single point on the value.  

```{r}
#work with smaller subset of data from economics, part of ggplot2 package 
startdate <- "2014-06-01"
economics_small <- economics %>%
  filter(date >= as.Date(startdate)) %>%
  arrange(date)
```


```{r}
A <- ggplot(economics_small, aes(date, unemploy))+
  geom_bar(stat="identity")+
  geom_vline(xintercept = as.Date(startdate), color="red", size=2)

B <- ggplot(economics_small, aes(date, uempmed))+
  geom_point()+geom_line()+
  geom_vline(xintercept = as.Date(startdate), color="red", size=2)

gA <- ggplotGrob(A) 
gB <- ggplotGrob(B)
grid::grid.draw(rbind(gA, gB)) #cowplot::plot_grid(A, B, ncol = 1, align = "v") produces same result 
```

In order to line the up there a a couple of options.  

#### Fix xlim for all charts 

If you make the limit the first x-value, the bar chart will not show up (remember it's centered over the value).  
```{r}
A <- ggplot(economics_small, aes(date, unemploy))+
  geom_bar(stat="identity")+
  geom_vline(xintercept = as.Date(startdate), color="red", size=2)+
  xlim(as.Date(startdate), NA)

B <- ggplot(economics_small, aes(date, uempmed))+
  geom_point()+geom_line()+
  geom_vline(xintercept = as.Date(startdate), color="red", size=2)+
  xlim(as.Date(startdate), NA)

gA <- ggplotGrob(A) 
gB <- ggplotGrob(B)
grid::grid.draw(rbind(gA, gB))
```

This can be fixed by adding a half unit to the x-axis (i.e. having the lower limit be half-unit lower than smallest x-value). In this case the unit is a month, so a half-unit would be ~15 days.  

```{r}
HalfUnit <- .5*(economics_small$date[2] - economics_small$date[1])
HalfUnit
```


```{r}
A <- ggplot(economics_small, aes(date, unemploy))+
  geom_bar(stat="identity")+
  geom_vline(xintercept = as.Date(startdate), color="red", size=2)+
  xlim(as.Date(startdate)-HalfUnit, NA)

B <- ggplot(economics_small, aes(date, uempmed))+
  geom_point()+geom_line()+
  geom_vline(xintercept = as.Date(startdate), color="red", size=2)+
  xlim(as.Date(startdate)-HalfUnit, NA)

gA <- ggplotGrob(A) 
gB <- ggplotGrob(B)
grid::grid.draw(rbind(gA, gB))
```

#### Shift Bar chart to right  
Bar charts are automatically centered over the x-value.  Bar charts (and any geom object) can be shifted by using `position - position_nudge())`.  The shift needs to be half a unit on the x-axis, again here it is monthly data so a half unit would be ~15 days.  
<a href="https://stackoverflow.com/questions/38752968/shift-geom-bar-right-not-center-aligned" target="_blank"> Source </a>

```{r}
A <- ggplot(economics_small, aes(date, unemploy))+
  geom_bar(stat="identity", position = position_nudge(x = as.vector(HalfUnit)))+
  geom_vline(xintercept = as.Date(startdate), color="red", size=2)

B <- ggplot(economics_small, aes(date, uempmed))+
  geom_point()+geom_line()+
  geom_vline(xintercept = as.Date(startdate), color="red", size=2)

gA <- ggplotGrob(A) 
gB <- ggplotGrob(B)

grid::grid.draw(rbind(gA, gB))
```

